\Large\textbf{Preparation GUIDE for Initial-Submission Manuscript}\\~\\ \large\textbf{ Preparation of Papers in Two-Column Format\\ for the ASP-DAC 2014 (\LaTeX2e version)}

### 2014-01-27 discussion.tex
Updated the paper

\section{Discussion -- addressing perceived limitations}
\label{sec:disc-perc-limit}

We dedicate this section to discuss the details of the delay semantics,
the dependence on time analyzable platforms, and the various uses and
programming paradigms that can be built using the \texttt{delay (M..N)}
construct.

\subsection{Programming using the delay construct}
\label{sec:progr-using-delay}

All three delay constructs are useful for programming real systems. % The
% first form of the delay construct $delay(M..N)$ is used to program
% non-determinism like in the HRCTS system and a real printer-spooler
% embedded system, etc. The second form is useful for programming low
% priority periodic tasks.

\subsubsection{Non-deterministic time}
\label{sec:non-determ-time}

Many real-world systems require non-deterministic timing
constructs. Where the exact real-time delay is not known or should not
be known apriori. One such system is our motivating example -- the human
response time system (HRTCS). Another one we presented was a real
printer-spooler embedded controller in
Section~\ref{sec:experimental-results}.

\subsubsection{Timeout}
\label{sec:timeout}

There are many instances when one would like to wait on an input from
environment for only a specified amount of time. This can be programmed
as:

\begin{figure}[h!]
  \centering
			\vspace{-10pt}
			\begin{lstlisting}[style=sysj,basicstyle=\normalsize\ttfamily,morekeywords={await,emit,trap,pause,exit,delay}]
// timeout after 1ms.
trap(T){{await(A);}||
        {delay(1 ms);exit(T);}}
\end{lstlisting}
  \caption{Programming low priority timeout tasks with delay statements}
  \label{fig:timeout}
			\vspace{-10pt}
\end{figure}

In the above case, the program waits for signal \texttt{A} from the
environment for \textit{at least} \texttt{1 ms}, but it may wait longer
if other reactions/CDs are scheduled for execution. This second variant
of delay statement is very useful for programming low priority tasks.

\subsubsection{Periodic reactions}
\label{sec:periodic-reactions}

The third variant is useful for hard-real time guarantees, e.g., a
reaction, or a whole CD, can be programmed to run periodically like so:

\begin{figure}[h!]
  \centering
	\vspace{-10pt}
		\begin{lstlisting}[style=sysj,basicstyle=\normalsize\ttfamily,morekeywords={emit,trap,pause,exit,delay}]
while(true) { delay ((1..1)ms); emit S; 
 //do something }
 \end{lstlisting}
  \caption{Programming periodic tasks with delay statements}
  \label{fig:periodic}
	\vspace{-10pt}
\end{figure}

The above program will emit signal \texttt{S} every \texttt{1 ms}. A
periodic reaction (or CD) requires special consideration, like any other
real-time periodic task. One should be aware of the interaction of
delays and other \texttt{pause} constructs in the body of the reaction
(or CD). For example, consider the (perceived) periodic program and its
rewrite below.

\begin{figure}[h!]
  \centering
  \begin{SubFloat}{\label{pp:a}Original program}
    \centering
			\begin{lstlisting}[style=sysj,basicstyle=\normalsize\ttfamily,morekeywords={emit,trap,pause,exit,delay}]
while(true) {
 delay(1..1 ms);
 //extra pause
 pause;
 emit S;
}
\end{lstlisting}
\end{SubFloat}
\hspace{9pt}
  \begin{SubFloat}{\label{pp:b}Rewritten program}
			\begin{lstlisting}[style=sysj,basicstyle=\normalsize\ttfamily,morekeywords={emit,trap,pause,exit,delay}]
while(true) {
 while(true) { 
  trap(T) {
   int x = 0;
   while(true) {
    x = x + 1;
    pause;
    if (x == d)
     exit(T);
   }
  }
  //extra pause
  pause;
  emit S;
 }
}
\end{lstlisting}
  \end{SubFloat}
\caption{Extra pause statements in periodic tasks}
  \label{fig:periodic2}
\end{figure}

The system designer needs to be aware that just introducing a delay
construct does not make the reaction (or CD) periodic. The program on
the left does \textit{not} emit signal \texttt{S} every \texttt{1 ms},
because of the extra \texttt{pause} statement. Introducing additional
pauses introduces more logical ticks, consequently increasing the
real-time delay. Static analysis techniques from real-time community can
be used to automatically determine the delay specification in such
cases, but describing these techniques is out of the scope of this work.

\subsection{Need for a time analyzable platform -- interaction of
  preemption and delays}
\label{sec:inter-preempt-delays}

@@

Preemption plays an important role in reactive languages. One needs to
carefully consider the interplay of \texttt{delay} semantics with the
preemption semantics of reactive languages. Previous attempts at
incorporating delays (using external timers) have only been partially
successful, because of the complex interplay between real-time and
preemption. Consider the simple example below, which models real-time
using external timers as in~\cite{rsh94}.

@@

As identified in~\cite{Bourke2009a} \texttt{suspend} does not play well
with the external timer. The above program sends a signal to an external
timer and waits for 10 ms to pass by. Consider what happens when signals
\texttt{S} and \texttt{TIMER} occur in the same logical tick, the
\texttt{await} statement is never executed (due to suspend) and hence,
we enter a deadlock. Such problems are completely avoided in our
technique, because we convert the real-time delays into logical delays
(\texttt{pause} constructs), which interact well with preemption.
Furthermore, synchronous/GALS languages have always been targeted at
real-time analyzable platforms and constrained
environments~\cite{DBLP:journals/pieee/SifakisTY03,boldt07}, one can use
our technique within such a setting with ease.

@@

\subsection{Interaction of channel communication and delays}
\label{sec:inter-chann-comm}

Channels, used for communication between reactions in asynchronously
running CDs, are an addition in the SystemJ language. Like interaction
of preemption and delays, conversion of real-time delays to logical
ticks also interacts well with channel rendezvous, because the semantics
of interaction are well defined~\cite{amal10}. More importantly, we need
to consider the interplay of channel communication and WCRT/BCRT
analysis. Since, channel communication does not stop logical time
(see~\cite{amal10}) WCRT/BCRT are unaffected by channel
communication. % The response time to input signals \textit{is} though!
% But, we are unconcerned with the response time, analysis in this paper
% and it remains a future research avenue.

>>>>

\section{Discussion}
\label{sec:disc-perc-limit}

We dedicate this section to discuss the details of the wait construct
semantics, their usage and the advantages and limitations of our
proposed technique to introducing these real-time constructs in a
reactive setting.

\subsection{Programming using the wait\_exactly construct -- the
  periodic reactions}
\label{sec:progr-using-delay}

% All three delay constructs are useful for programming real systems. % The
% first form of the delay construct $delay(M..N)$ is used to program
% non-determinism like in the HRCTS system and a real printer-spooler
% embedded system, etc. The second form is useful for programming low
% priority periodic tasks.

% \subsubsection{Non-deterministic time}
% \label{sec:non-determ-time}

% Many real-world systems require non-deterministic timing
% constructs. Where the exact real-time delay is not known or should not
% be known apriori. One such system is our motivating example -- the human
% response time system (HRTCS). Another one we presented was a real
% printer-spooler embedded controller in
% Section~\ref{sec:experimental-results}.

% \subsubsection{Timeout}
% \label{sec:timeout}

% There are many instances when one would like to wait on an input from
% environment for only a specified amount of time. This can be programmed
% as:

% \begin{figure}[h!]
%   \centering
% 			\vspace{-10pt}
% 			\begin{lstlisting}[style=sysj,basicstyle=\normalsize\ttfamily,morekeywords={await,emit,trap,pause,exit,delay}]
% // timeout after 1ms.
% trap(T){{await(A);}||
%         {delay(1 ms);exit(T);}}
% \end{lstlisting}
%   \caption{Programming low priority timeout tasks with delay statements}
%   \label{fig:timeout}
% 			\vspace{-10pt}
% \end{figure}

% In the above case, the program waits for signal \texttt{A} from the
% environment for \textit{at least} \texttt{1 ms}, but it may wait longer
% if other reactions/CDs are scheduled for execution. This second variant
% of delay statement is very useful for programming low priority tasks.

% \subsubsection{Periodic reactions}
% \label{sec:periodic-reactions}

% The third variant is useful for hard-real time guarantees, e.g., a
% reaction, or a whole CD, can be programmed to run periodically like so:

\begin{figure}[h!]
  \centering
	\vspace{-10pt}
        \begin{lstlisting}[style=sysj,basicstyle=\normalsize\ttfamily,morekeywords={emit,trap,pause,exit,wait_exactly}]
          while(true){wait_exactly (1 ms); emit S; //do something
          }
 \end{lstlisting}
  \caption{Programming periodic tasks with wait\_exactly statement}
  \label{fig:periodic}
	\vspace{-10pt}
\end{figure}

The program in Figure~\ref{fig:preemp}, reproduced from
Section~\ref{sec:progr-using-exact}, is supposed to emit signal
\texttt{S} every \texttt{1 ms}. Such periodic reactions (or CDs) require
special consideration, like any other real-time periodic task. One
should be aware of the interaction of wait and other \texttt{pause}
constructs in the body of the reaction (or CD). For example, consider
the (perceived) periodic program and its rewrite below.

\begin{figure}[h!]
  \centering
  \begin{SubFloat}{\label{pp:a}Original program}
    \centering
			\begin{lstlisting}[style=sysj,basicstyle=\normalsize\ttfamily,morekeywords={emit,trap,pause,exit,wait_exactly}]
while(true) {
 wait_exactly(1 ms);
 //extra pause
 pause;
 emit S;
}
\end{lstlisting}
\end{SubFloat}
\hspace{9pt}
  \begin{SubFloat}{\label{pp:b}Rewritten program}
			\begin{lstlisting}[style=sysj,basicstyle=\normalsize\ttfamily,morekeywords={emit,trap,pause,exit,delay}]
while(true) {
 while(true) { 
  trap(T) {
   int x = 0;
   while(true) {
    x = x + 1;
    pause;
    if (x == d)
     exit(T);
   }
  }
  //extra pause
  pause;
  emit S;
 }
}
\end{lstlisting}
  \end{SubFloat}
\caption{Extra pause statements in periodic tasks}
  \label{fig:periodic2}
\end{figure}

The system designer needs to be aware that just introducing a wait
construct does not make the reaction (or CD) periodic. The program on
the left does \textit{not} emit signal \texttt{S} every \texttt{1 ms},
because of the extra \texttt{pause} statement. Introducing additional
pauses introduces more logical ticks, consequently increasing the
real-time wait constructs. Static analysis techniques from real-time
community can be used to automatically determine the wait specification,
in such cases, but describing these techniques is out of the scope of
this work.

\subsection{Need for time analyzable platforms}
\label{sec:need-time-analyzable}

The proposed solution relies on time analyzable platforms, i.e.,
platforms where WCRT and BCRT can be calculated using static analysis
tools. Many might consider this to be too much of a restriction. In this
section we debunk this \textit{perceived} disadvantage.

\subsubsection{Interaction of timers and reactive constructs}
\label{sec:inter-timers-react}

@@

Preemption plays an important role in reactive languages. One needs to
carefully consider the interplay of wait constructs semantics with the
preemption semantics of reactive languages. Previous attempts at
incorporating delay (using external timers) have only been partially
successful, because of the complex interplay between real-time and
preemption. Consider the simple example below, which models real-time
using external timers as in~\cite{rsh94}.

@@

As identified in~\cite{Bourke2009a} \texttt{suspend} does not play well
with the external timer. The above program sends a signal to an external
timer and waits for 10 ms to pass by. Consider what happens when signals
\texttt{S} and \texttt{TIMER} occur in the same logical tick, the
\texttt{await} statement is never executed (due to suspend) and hence,
we enter a deadlock. Such problems are completely avoided in our
technique, because we convert the real-time wait constructs into logical
waits (\texttt{pause} constructs), which interact well with preemption.
Furthermore, synchronous/GALS languages have always been targeted at
real-time analyzable platforms and constrained
environments~\cite{DBLP:journals/pieee/SifakisTY03,boldt07}. One can use
our technique within such a setting with ease.

\subsubsection{The timer resolution problem}
\label{sec:resolution-real-time}

\begin{figure}[h!]
  \centering
	\vspace{-10pt}
		\begin{lstlisting}[style=sysj,basicstyle=\normalsize\ttfamily,morekeywords={emit,trap,pause,exit,delay,suspend}]
{
 //reaction R1
 emit START_TIMER(1 ms); 
 await (TIMER);emit O1;
}
||
{
 //reaction R2
 // do something
}
		\end{lstlisting}
  \caption{Resolution of external timers}
  \label{fig:resolution}
	\vspace{-10pt}
\end{figure}

There is yet another problem with timer based systems. Let us consider
the program in Figure~\ref{fig:resolution}. On a cursory look this
example should work fine. An external timer is started in reaction R1,
which counts down from 1 ms. Once this time has elapsed a signal
\texttt{TIMER} is generated from this external timer, which should emit
O1 in turn. There is no \texttt{suspend} construct encapsulating the
timer and hence all seems fine. Let us now consider what happens to the
whole CD, including reaction R2; while reaction R1 is waiting for the
\texttt{TIMER} signal, reaction R2 is making progress. Input signals can
only be captured from the environment at the logical tick boundaries
(cf. Section~\ref{sec:mapping-logical-time}). In
Figure~\ref{fig:resolution}, reaction R2 determines the length of the
tick, because it is performing the heavier computation. Consider what
happens if the external timer generates the \texttt{TIMER} signal, but
the system is still performing computation and the logical tick has not
yet finished -- we \textit{miss} this \texttt{TIMER} signal from the
environment. Thus, even in the absence of the \texttt{suspend}
construct, we are not guaranteed that the program will capture the
external timer signal. In fact, the WCRT needs to be smaller than 1 ms
for the above program to perform as expected. But, this again implies
need for a time analyzable platform even when using external timers.

In general we observe that the WCRT determines the lowest resolution of
any real-time construct whether it be external timer independent, like
the proposed technique, or external timer dependent, like above. This
discovery is one of the key insights for proposing the solution
described in this paper.

@@

\subsection{Interaction of channel communication and wait constructs}
\label{sec:inter-chann-comm}

Channels, used for communication between reactions in asynchronously
running CDs, are an addition in the SystemJ language. Like interaction
of preemption and waits, conversion of real-time wait constructs to
logical ticks also interacts well with channel rendezvous, because the
semantics of interaction are well defined~\cite{amal10}. More
importantly, we need to consider the interplay of channel communication
and WCRT/BCRT analysis. Since, channel communication does not stop
logical time (see~\cite{amal10}) WCRT/BCRT are unaffected by channel
communication. % The response time to input signals \textit{is} though!
% But, we are unconcerned with the response time, analysis in this paper
% and it remains a future research avenue.
